---
layout: post
title: Spotify + R = <3
author: tad
tags:
- R
- Spotify
---


If I'm in the lab, I most likely have my headphones on. There is [documented evidence of this](https://github.com/taddallas). I remember when Spotify Apps first became a thing, and I thought it was really cool that Spotify had an API, so you could query related artists and such pretty easily. This post is about some fun Spotify querying in R!


## The problem

I initially got excited about the idea of the "Nickelback number", which would be a Spotify App that calculate the degrees of separation between an input artist and the Canadian rock band [Nickelback](http://en.wikipedia.org/wiki/Nickelback). However, that turned out to require a bit too much work, and it felt a little mean towards the Canadian **pop** rockers. Instead, the point of this blog post is to show how to access Spotify's API (Application Program Interface) for some fun!


## Why is this problem non-trivial?

It is pretty trivial, to be honest. However, learning how to query API's may be generally useful to ecologists. If you need examples, check out some of the great packages at [rOpenSci](http://ropensci.org/)! 


## What I did

All code is available as part of the `rSpotify` package available on my [Github](http://github.com/taddallas). The package is bare bones, but has some cool plotting functions using the `d3Network` package. 

Information on artists can be obtained by using the following base URL.

```{r eval=FALSE, echo=TRUE}
https://api.spotify.com/v1/artists/ARTISTID
```

```{r eval=TRUE, echo=FALSE, message=FALSE}
library(RCurl); library(jsonlite);
```

ARTISTID is the Spotify code for an artist. You can access that code using the `RCurl` and `jsonlite` packages. Below, I access the API using the `getURL` function, and fromJSON parses JSON objects into R. 

```{r eval=TRUE, echo=TRUE}
	artistName="The Mountain Goats"

  info=fromJSON(getURL(paste("https://api.spotify.com/v1/search?q=", gsub(' ', '+', artistName),"&type=artist", sep='')))$artists$items
 
  name=info$name[1]
  id=unlist(strsplit(info$uri[1],':'))[3]
```

Now we have the artists name and their Spotify URI, a unique artist identifier. From this, we can get information about how many followers they have, their popularity score, band image urls, and related artists.

```{r eval=TRUE, echo=TRUE}
 site=paste("https://api.spotify.com/v1/artists/",id,"/related-artists", sep='')
  
  bandsInfo=as.data.frame(fromJSON(getURL(site, httpheader = c(Accept = "application/json"))))
```

This `bandsInfo` object contains information on 20 artists related to the original query ("The Mountain Goats"). Using this tiny dataset, we can explore relationships like the one between the number of artist followers, and the popularity score (which is calculated based on the number of plays per some unit time). 



```{r eval=TRUE, echo=TRUE}
plot(unlist(bandsInfo[,2])[21:40], bandsInfo[,8], ylab='Popularity', xlab='Number of followers', las=1, pch=16, tck=0.01)
```

Looks cool. Now let's look at how Nickelback compares. 

```{r eval=TRUE, echo=TRUE}
	awfulBand="Nickelback"

  awfulBandInfo=fromJSON(getURL(paste("https://api.spotify.com/v1/search?q=", gsub(' ', '+', awfulBand),"&type=artist", sep='')))$artists
 
  awfulPop=awfulBandInfo$items[1,8]
 	awfulFollowers=unlist(awfulBandInfo$items[,2])['total1']

	plot(unlist(bandsInfo[,2])[21:40], bandsInfo[,8], ylab='Popularity', xlab='Number of followers', las=1, pch=16, tck=0.01, xlim=c(15000,1100000), ylim=c(45, 90))

	points(awfulFollowers, awfulPop, pch=1, col='red', cex=2)
	text(1004000, 80, "Nickelback!", col='red', cex=1.5)
```

Looks like Nickelback dominates The Mountain Goats, and all of their related artists, including Athens favorites Of Montreal! Sad day, but it's understandable. I hope this has been fun for you. To conclude the post, and put things right with the universe, I'd like to show Katy Perry as a dark horse, destroying Nickelback in popularity and followers. 


```{r eval=TRUE, echo=TRUE}
	
  KP=fromJSON(getURL(paste("https://api.spotify.com/v1/search?q=", gsub(' ', '+', "Katy Perry"),"&type=artist", sep='')))$artists
 
  kpPop=KP$items[1,8]
 	kpFollowers=unlist(KP$items[,2])['total1']

	plot(unlist(bandsInfo[,2])[21:40], bandsInfo[,8], ylab='Popularity', xlab='Number of followers', las=1, pch=16, tck=0.01, xlim=c(15000,3500000), ylim=c(45, 90))

	points(awfulFollowers, awfulPop, pch=1, col='red', cex=2)
	text(1004000, 80, "Nickelback!", col='red', cex=1.5)

	points(kpFollowers, kpPop, pch=1, col='dodgerblue2', cex=2)
	text(3000000, 85, "Katy Perry", col='dodgerblue2', cex=1.5)
```

This has been a bit of a roundabout way to show how to access Spotify's API through R, but I hope it has got you thinking about accessing data from the web through API's. 












