---
layout: post
title: Detecting Allee effects with maximum likelihood estimation
author: reni
tags:
- parameter estimates
- mle
---


## The Concept: Allee effect
Many mechanisms impact the survival of a population. One such mechanism is an Allee effect- or postive density dependence. Simply put, populations grow slower when their aren't enough individuals in the population. Perfect examples are hunting packs where a minimum number is needed to capture prey, or schooling fish which reduce their individual mortality rate by clumping together. Of course the most obvious mechanism leading to an Allee effect is sexual reproduction. A sexually reproducing population needs at least a female and male in a density that they are likely to find eachother after an appropriate amount of searching. 
 
Allee effects come in two flavors, weak and strong. A weak effect will just slow population growth until a specific density is reached; a small school of fish will grow slower since more individuals will be eaten than if it was larger.  On the other hand, a strong Allee effect will results in a negative  growth rate; a population consisting of only males will slowly decline until the populations goes extinct. For more information check out [this article](http://www.nature.com/scitable/knowledge/library/allee-effects-19699394) written by John Drake and Drew Kramer in *Nature Education Knowledge*. 

## The Question: Is a population subject to an Allee effect?
As you can imagine, the presence and strength of an Allee effect will change how endangered or nuisance populations would be managed. For this reason, it is important to test for the presence of Allee effects. A strong Allee effect is present if the relationship between probability of establishment and starting population density is sigmoidal. We will fit the experimental survivorship data to a Weibull function.

```{r, echo=FALSE}
#plot simulated data, log transformed with fitted line
par(mar=c(8,3,1,1))
x<-seq(0,100, by=1)
plot(NA, xlim=c(0,100), ylim=c(0,1),xaxt='n', xlab='Initial Population', ylab='Probability of Establishment')
axis(1, at=c(0,25,50,75,100))
legend(x=36, y=0.5, legend=c('Strong Allee effect','No Allee effect'), lty=1, lwd=3, col=c(1,2), cex=1.5)
#plot(NA, ylim=c(0,1), xlim=c(0,250))
#plot weibull line with given parameters

#par.series= k,l
par.series=matrix(c(.8,1, #black
                    2,40), #red
                  nrow=2)

for (i in 1:length(par.series[1,])){
  k<- par.series[1,i]
  l<- par.series[2,i]
  lgy<-1-exp(-(x/l)^k)
lines(lgy~x, col=i, lwd=3) 
}
#crappy caption
mtext('Figure 1. Caption here explaining ', side=1, line=4.5, outer=FALSE, adj=0)
```


## The Generic Question: How can I fit binary data to a Weibull function?

The Weibull function can be used to test how a function reaches the upper bound. Is the approach sigmodial or ?

\Weibull fitting and MLE useful in other situations 

## What I did

      
```{r echo=TRUE, eval=TRUE}
```





