---
layout: post
title: Animating the double well potential
author: john
tags:
- data visualization
- animation
- dynamical systems
---

I recently found myself wanting an animation of fluctuations in a double well potential to use in a presentation. The double well potential is a mathematical object from statistical physics -- the classical "ball-and-cups" diagram -- but it's useful for thinking about bistable systems in general. It turns out, this was quite easy to do using the `animation` package in R.

To start, I searched for a simple formula to serve as the potential function and discovered the [Duffing equation](http://physics.ucsc.edu/~peter/115/duffing.pdf), which was previously unfamiliar to me. The equation, $V(x) = -x^2/2 + x^4/4$, is plotted below using the following code snippet.

```{r duffing}
duffing <- function(x) -x^2/2 + x^4/4
xi <- seq(-2,2, length=100)
y <- duffing(xi)
plot(xi, y, type='l', lwd=2, xlab='x', ylab='V(x)')
```

For my illustration, I wanted the potential barrier (the hump in the middle) to be higher and for the potential function itself to be asymmetric. I modified the original equation by multiplying the first term by a constant and adding a third term that is linear in $x$, yielding $W(x) = -2(x^2/2) + x^4/4 + ax = - x^2 + x^4/4 + ax$. My function, with $a=0.1$, looks like this.

```{r potential}
W <- function(x, a=0) -2*(x^2/2) + x^4/4 + a*x
xi <- seq(-2,2, length=100)
y <-W(xi, a=0.1)
plot(xi, y, type='l', lwd=2, xlab='x', ylab='W(x)')
```

To simulate the dynamics of a particle subject to this potential we need to differentiate $W$ and take the negative, yielding $w = 2x - x^3 -a$.

```{r derivative}
w <- function(x, a=0) 2*x - x^3 - a
```

Next we write a function to solve for the fluctuations using [Euler's method](http://en.wikipedia.org/wiki/Euler%E2%80%93Maruyama_method) for [stochastic differential equations](http://en.wikipedia.org/wiki/Stochastic_differential_equation). Here the standard deviation is arbitrarily set to 0.5 and the time scale is determined by setting $\tau=1$.

```{r euler}
euler <- function(x0, a=0, tau=1, sigma=0.5){
  return(x0 + w(x0, a)*tau + sigma*rnorm(1, mean=0, sd=sqrt(tau)))
}
```

Now we write a function that can produce an animation through repeated plotting. As written, the function doesn't require any arguments (relevant constants are set in the first five lines of the function). A loop is used to produce 1000 plots, each of which consists of two panels. The top panel is a picture of the potential function with the position of the particle indicated by a point. The bottom panel is a time series showing the last 100 positions. Time is shown on the $y$-axis so that the position of the particle in the two graphs can be aligned.

```{r animate-fxn}
animate <- function(){
  a <- 0.1
  xi <- seq(-2,2, length=100)
  y <-W(xi, a=a)
  x <- 1
  t <- 0
  for(i in 1:1000){
    layout(matrix(c(1,2,2,2), nrow=4, ncol=1, byrow = TRUE))
    par(mar=c(1,4,2,2))
    plot(xi,W(xi, a=a), type='l', lwd=4, xlab='', ylab='', axes=FALSE)
    box()
    
    t[i+1] <- i
    x[i+1] <- euler(tail(x,1), a=a, tau=0.1)
    points(x[i+1], W(x[i+1],a)+0.1, cex=3, pch=21, bg='red')
    
    par(mar=c(5,4,0,2))
    plot(tail(x,100), tail(t,100), type='l', xlab='', ylab='Time', col='red', xlim=c(-2,2))
  }
}
```

Finally, this function is called within a code chunk using the `animation` package to produce the individual frames (as bitmap images) and stitch them together.

```{r animation, echo=FALSE, fig.show='animate', aniopts='controls, loop', interval=0.1, nmax=1001, ani.width=480, ani.height=480}
par(bg='white')
library(animation)
animate()
```
